---
title: "Traffic Stops"
---
The following data analysis uses data compiled from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/), which works to create "a comprehensive, national repository detailing interactions between police and the public." The database includes standardized data on traffic stops (vehiclular and pedestrian) from law enforcement departments from 42 states.

One area of interest in the data is regarding vehicular searches. Approximately half of the datasets include data on whether or not a search was conducted. Vehicular searches are often prompted by suspicion of possession of illicit drugs. My analysis focuses on this question: Does the legalization of marijuana affect vehicle search rates?

I expect to see a decline in search rates after the legalization of marijuana as police officers no longer need to conduct vehicular searches if a person possesses marijuana since it is not illegal.

We begin by setting up the SQL connection.

```{r message = FALSE, warning = FALSE}
#| echo: false
library(RMariaDB)
library(tidyverse)
con_traffic <- DBI::dbConnect(

  RMariaDB::MariaDB(),

  dbname = "traffic",

  host = Sys.getenv("TRAFFIC_HOST"),

  user = Sys.getenv("TRAFFIC_USER"),

  password = Sys.getenv("TRAFFIC_PWD")

)
```

Of the states included in the dataset, three states (California, Colorado, Washington) have considerable data available from both before and after their legalization of marijuana. My analysis focuses on these three states, since data from before and after legalization makes the effect of legalization visible.


```{sql}
#| connection: con_traffic
#| output.var: "CO_table"
SELECT COUNT(*) AS num_stops,
  SUM(search_conducted) AS num_searched,
  AVG(search_conducted)*100 AS percent_searched,
  date >= "2012-12-10" AS after_legal
FROM co_statewide_2020_04_01
GROUP BY after_legal
HAVING (after_legal IS NOT NULL)
```


```{sql}
#| connection: con_traffic
#| output.var: "WA_table"
SELECT COUNT(*) AS num_stops,
  SUM(search_conducted) AS num_searched,
  AVG(search_conducted)*100 AS percent_searched,
  date >= "2012-12-06" AS after_legal
FROM wa_statewide_2020_04_01
GROUP BY after_legal
HAVING (after_legal IS NOT NULL)
```


```{sql}
#| connection: con_traffic
#| output.var: "CA_table"
SELECT
    SUM(num_stops) AS num_stops,
    SUM(num_searched) AS num_searched,
    SUM(num_searched)*100/SUM(num_stops) AS percent_searched,
    (date >= '2016-11-09') AS after_legal
FROM (SELECT 1 AS num_stops,
      date,
      search_conducted AS num_searched
    FROM ca_san_diego_2020_04_01
    
    UNION ALL

    SELECT 1,
      date,
      search_conducted
    FROM ca_oakland_2020_04_01
    
    UNION ALL
    
    SELECT 1,
      date,
      search_conducted
    FROM ca_san_jose_2020_04_01
    
    UNION ALL
    
    SELECT 1,
      date,
      search_conducted
    FROM ca_stockton_2020_04_01
    
    ) AS cali_table
GROUP BY after_legal
HAVING (after_legal IS NOT NULL);
```

The following tables summarize the results. The num_stops column indicates the number of traffic stops, the num_searched column indicates the number of stops in which a search was conudcted, and the percent_searched column indicates the percentage of stops in which a seach was conducted (the 0.4899 below indicates 0.4899%, not 48.99%). Lastly, the after_legal column indicates whether that row refers to traffic stops before or after the legalization of marijuana in that state.

```{r}
CO_table <- CO_table |> mutate(after_legal = factor(after_legal, levels = c(0, 1), labels = c("Before", "After")))
WA_table <- WA_table |> mutate(after_legal = factor(after_legal, levels = c(0, 1), labels = c("Before", "After")))
CA_table <- CA_table |> mutate(after_legal = factor(after_legal, levels = c(0, 1), labels = c("Before", "After")))

print(CO_table)
print(WA_table)
print(CA_table)
```

The below graphs show the impact of  We can see that Colorado and Washington saw nearly 50% decreases in vehicular search rates after legalization of marijuana. However, California saw a 50% increase in vehicular searches.

```{r}
ggplot(CO_table,
       aes(x = after_legal,
           y = percent_searched)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous() +
  labs(title   = "Search Rates Before vs. After Legalisation in Colorado",
       x = "",
       y       = "% of Stops that Led to a Search",
       fill    = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

ggplot(WA_table,
       aes(x = after_legal,
           y = percent_searched)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous() +
  labs(title   = "Search Rates Before vs. After Legalisation in Washington",
       x       = "",
       y       = "% of Stops that Led to a Search",
       fill    = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

ggplot(CA_table,
       aes(x = after_legal,
           y = percent_searched)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous() +
  labs(title   = "Search Rates Before vs. After Legalisation in California",
       x       = "",
       y       = "% of Stops that Led to a Search",
       fill    = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")

total_table <- CO_table |> mutate(state = "CO") |> rbind(mutate(WA_table, state = "WA")) |> rbind(mutate(CA_table, state = "CA")) |> group_by(after_legal) |> summarize(percent_searched = sum(num_searched)/sum(num_stops))

ggplot(total_table,
       aes(x = after_legal,
           y = percent_searched)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_y_continuous() +
  labs(title   = "Search Rates Before vs. After Legalisation in CO, WA, and CA",
       x       = "",
       y       = "% of Stops that Led to a Search",
       fill    = "") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```


The data from Colorado and Washington were taken from state patrol on highways, while the data from California was complied from police departments within cities, which could help explain the discrepancy in the number of traffic stops recorded in CA vs. WA and CO. The Washington and Colorado traffic stops outnumbered the California stops to such a degree that there is an overall decrease in search rates after legalization, despite California's large percentage increase in search rates.

Overall, it is difficult to determine from the data if marijuana legalization led to decreased search rates, although search rates did decrease significantly in Washington and Colorado. The California analysis would benefit from statewide data that would make it more comparable to the Washington and Colorado data, as opposed to the city-specific data (and relatively low number of obersvations) available for California.

One limitation of this analysis is that it doesn't recognize the possible changes in searches due to other reasons, such as falling crime or changes in policing practices. In addition, since data is limited to only certain cities in California, while Colorado and Washington use state patrol data, different policing practices among these groups could create greater discrepancies. Further areas for exploration would be to compare the above results to states which have not legalized recreational marijuana to isolate the impact of policy on traffic searches, along with how search rates vary across groups by race.

References:

Pierson, Emma, Camelia Simoiu, Jan Overgoor, Sam Corbett-Davies, Daniel Jenson, Amy Shoemaker, Vignesh Ramachandran, et al. 2020. “A Large-Scale Analysis of Racial Disparities in Police Stops Across the United States.” *Nature Human Behaviour*, 1–10.
